Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets

### 2021-07-10 introduction.tex
Révision intro

% !TEX root = main.tex
%
% Introduction
%
Parsing is the problem
of structuring a linear representation
(a finite word) according to a language model. % (a formal grammar).
% natural language, programming language,
%
Most of the context-free parsing approaches~\cite{GruneJacobs08parsing}
assume a finite and reasonably small input alphabet. %models and algorithms
Such a restriction makes perfect sense in the context of
NLP tasks such as constituency parsing,
or of programming languages compilers or interpreters.
Considering large or infinite alphabets can however be of
practical interest, when dealing with large characters encodings such as UTF-16,
% processing strings in \eg for vulnerability detection in Web-applications~\cite{dAntoni21CACM},
%
analysis of data streams or serialization of structured documents~\cite{Segoufin06csl},
or processing timed execution traces~\cite{Bouyer03algebraic}.
% algebraic definition of a class of data languages
% (notion of monoid recognizability, based on registers, comparable to Bojancszik et al. data words)
%

The latter case is related to a motivation  of  the present work:
automated music transcription. Representations that capture  music performances
are
essentially linear:   audio files, or the widely used MIDI format~\cite{?}.
Such representations ignore the hierarchical structures that frame the
conception of music, at least in the western area. These structures, on the other hand,
are present, either explicitly  or implicitly,
in music notation~\cite{Gould11Notation}: music scores are partitioned in measures, measures
in beats, and beats can be further recursively divided.
It follows that music events do not occur at arbitrary timestamps,
but respect a discrete partitioning of the  timeline incurred by
these recursive divisions.
The \emph{transcription problem} takes
as input a linear representation (audio or MIDI) and aims at re-constructing
these structures
by mapping input events to this hierarchical rhythmic space.
It can therefore be stated as a parsing problem~\cite{foscarin:hal-01988990},
over an infinite alphabet of timed events.

Various extensions of language models for handling infinite alphabets have been studied.
%words carrying data values in an infinite domain (e.g. integers) e.g. data processing
For instance, some automata with memory extensions
allow restricted storage and comparison of input symbols,
%and correspond with logics,
(see~\cite{Segoufin06csl} for a survey),
\florent{register: skip refs and details, add Mikolaj recent}
with pebbles for marking positions~\cite{NevenSchwentickVianu04FSMinfinite},
registers~\cite{KaminskiFrancez94},
or %computations in 2 steps
the possibility to compute on subsequences
with the same attribute values~\cite{Bojanczyk11FO2}. % data words automata.
%
%for the and verification of infinite state systems
%(model checkers: alphabet = long bit-vectors)
%...For the representation of  in model checking, verification and
%
The automata at the core of model checkers
compute on input symbols represented by large bitvectors~\cite{Vardi07ciaa} %\cite{BaierKatoen08MC}
(sets of assignments of Boolean variables) %propositional variables)
and in practice,  %implementation
each transition accepts a set of such symbols (instead of an individual symbol),
represented by Boolean formula or Binary Decision Diagrams.
%
Following a similar idea, % and generalizing,
in symbolic automata (\SA)~\cite{dAntoniVeanes17CAV,dAntoni21CACM},
the transitions are guarded by predicates over infinite alphabet domains.
With appropriate closure conditions on the sets of such predicates, % (alphabet theories),
all the good properties enjoyed by automata over finite alphabets are preserved.

Other extensions of language models  %(automata and grammars)
help in dealing with non-determinism, by the computation of weight values.
With an ambiguous grammar, there may exist several derivations
(\emph{abstract syntax trees} -- AST) % \ie the result of an analyze.
yielding one input word. % structuring a word,
The association of one weight value %in semiring domain
to each AST permits to select a best one (or $n$ bests). % a fixed number of bests
% = ...a ranking of ASTs
%This is roughly the principle of \emph{weighted parsing} approaches
\florent{WARNING: \cite{MorbitzVogler19weighted-parsing} much is more general that weighted parsing}
In \emph{weighted language models}~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing},
like \eg probabilistic context-free grammars % (CFG),
and weighted automata (\WA) \cite{Droste09handbook},
a weight value is associated to each transition rule, % production rule
and the rule's weights can be combined with an
associative product operator~$\otimes$ into the weight of an AST.
A second operator~$\oplus$, associative and commutative,
is moreover used to resolve the ambiguity raised by the existence
of several (in general exponentially many) AST
associated to a given input word.
Typically, $\oplus$ will select the best of two weight values.
%Intuitively,~$\oplus$ selects, or ranking, the syntax trees.
The weight domain, equipped with these two operators shall be, at minima,
a \emph{semiring} %$\Semiring$
where $\oplus$ can be extended to infinite sums,
%\cite{Eilenberg74automata}
such as the Viterbi semiring and the tropical min-plus algebra% % see Figure~\ref{fig:semirings}.
%of domain $\mathbb{R}^+ \cup \{ +\infty\}$,
%where $\oplus$ is min and $\otimes$ is plus .
%by ranking
%making the weight domain a semiring.
%Some efficient specialized parsing algorithms~\cite{Huang05kbest} have been proposed in this context
%% models represented as hypergraphs \cite{Huang05kbest}
%in order to compute the $n$ best syntax trees of a given input word without having to enumerate them all.
%Generally based on dynamic programming, these algorithms rely on
%additional algebraic properties of~$\Semiring$.
%-- see \eg~\cite{Huang05kbest} for some NLP applications.
%The extraction of $n$ best list is useful
%the problem: quantitative parsing or symbolic parsing
%of parsing of words over infinite input alphabet.

In this paper, we present a uniform framework for weighted parsing over infinite input alphabets.
It is based on \emph{symbolic weighted} finite states language models (\swM),
generalizing the Boolean guards of~\SA %in transitions
to functions into an arbitrary semiring,
and generalizing also~\WA, by handling infinite alphabets, see Figure~\ref{fig:hierarchy}.

%***
%some weighted language models
%model of weighted CFG
%computing on words over infinite input alphabets %sets of terminal symbol
%- but with finite sets of states and transitions rules.
% non-terminals and production rules).

@@

@@

@@

The latter model of automata operates sequentially on \emph{nested words}~\cite{AlurMadhusudan09nested},
structured with markup symbols (parentheses) and able to describe linearizations of trees.

@@

In the context of parsing, they can represent (weighted) AST of CF grammars.
More precisely, a \SWVPA $A$ associates a weight value $A(t)$ % \in \Semiring$
to a given nested word $t$, which is the linearization of an AST. %representing a parse tree.
On the other hand,
a \SWT can define a distance $T(s, t)$ between finite words $s$ and $t$
Then, the \emph{SW-parsing} problem aims at %computing

@@

$T(s, t) \otimes A(t)$ (\wrt the ranking defined by $\oplus$), given an input word $s$.
The latter product is called the distance between $s$ and $A$ in~\cite{Mohri03EDWA}.

@@

our approach proceeds in two steps,
based on properties of the \swM.

@@

\begin{example}[Running example]\label{ex:running}
Throughout the paper we
illustrate our framework with music transcription examples: Given
a \emph{timeline} of musical events with arbitrary timestamps from an infinite
domain as input, parse it
into a structured music score. In our example, input events
are pairs $\< \mu, \tau>$ made of
a symbol $\mu \in \Sigma$, where $\Sigma$ stands for the
set of MIDI message symbols~\cite{?}
and  $\tau \in \mathbb{Q}$ is a timestamp. Such inputs typically correspond
to the recording of a live performance. The output of parsing
is a sequence of events (or \emph{notes}) $\< \nu, \tau'>$   in
Common Western Music Notation (CWMN)~\cite{Gould11Notation}
where event symbols $\nu$ belong to the domain $\Delta$
of \emph{pitches} (e.g., A4, G5, etc.), and timestamp $\tau'$
belongs to a rhythmic ``grid'' obtained from recursive divisions:
whole notes ($\musWhole$) splitted in halves ($\musHalf$), halves
in quarters ($\musQuarter$), eights ($\musEighth$), etc.

The following concrete sequences
will be used:
\begin{enumerate}
  \item A performance  $I = [ \<\mu_1, 0.07>, \<\mu_2, 0.72>, \<\mu_3, 0.91>, \<\mu_4, 1.05>, \<\mu_5, 1.36>, \<\mu_6, 1.71>]$,
     over interval $[0,2[$,
  \item A score $O =$ \includegraphics[scale=0.20]{pictures/score5.png}, corresponding to
      a hierarchical structures (measures, tuplets) that can be linearized as a sequence of events
      $[\<\nu_1,0>, \<\nu_2,\frac{3}{4}>, \<\nu_3,\frac{7}{8}>, \<\nu_4,1>, \<\nu_5,\frac{4}{3}>, \<\nu_6\frac{5}{3}>]$
\end{enumerate}
\smallskip

We will show that $O$  can result for the
parsing of $I$. Note that several other parsings (e.g.,
\includegraphics[scale=0.20]{pictures/score4.png}) are possible.
SW-parsing associates a weight value
to each solution, and our framework
aims at selecting the best one with respect to this weight.
\endex
\end{example}

>>>>

% !TEX root = main.tex
%
% Introduction
%
Parsing is the problem
of structuring a linear representation
(a finite word) according to a language model. % (a formal grammar).
% natural language, programming language,
%
Most of the context-free parsing approaches~\cite{GruneJacobs08parsing}
assume a finite and reasonably small input alphabet. %models and algorithms
Such a restriction makes perfect sense in the context of
NLP tasks such as constituency parsing,
programming languages compilers or interpreters.
Considering large or infinite alphabets can however be of
practical interest, when dealing with large characters encodings such as UTF-16,
% processing strings in \eg for vulnerability detection in Web-applications~\cite{dAntoni21CACM},
%
analysis of data streams, serialization of structured documents~\cite{Segoufin06csl},
or processing timed execution traces~\cite{Bouyer03algebraic}.
% algebraic definition of a class of data languages
% (notion of monoid recognizability, based on registers, comparable to Bojancszik et al. data words)
%

The latter case is related to a motivation  of  the present work:
automated music transcription. Representations that capture  music performances
are
essentially linear:   audio files, or the widely used MIDI format~\cite{?}.
Such representations ignore the hierarchical structures that frame the
conception of music, at least in the western area. These structures, on the other hand,
are present, either explicitly  or implicitly,
in music notation~\cite{Gould11Notation}: music scores are partitioned in measures, measures
in beats, and beats can be further recursively divided.
It follows that music events do not occur at arbitrary timestamps,
but respect a discrete partitioning of the  timeline incurred by
these recursive divisions.
The \emph{transcription problem} takes
as input a linear representation (audio or MIDI) and aims at re-constructing
these structures
by mapping input events to this hierarchical rhythmic space.
It can therefore be stated as a parsing problem~\cite{foscarin:hal-01988990}
over an infinite alphabet of timed events.

Various extensions of language models for handling infinite alphabets have been studied.
%words carrying data values in an infinite domain (e.g. integers) e.g. data processing
Some automata with memory extensions
allow restricted storage and comparison of input symbols,
%and correspond with logics,
(see~\cite{Segoufin06csl} for a survey),
\florent{register: skip refs and details, add Mikolaj recent}
with pebbles for marking positions~\cite{NevenSchwentickVianu04FSMinfinite},
registers~\cite{KaminskiFrancez94},
or %computations in 2 steps
the possibility to compute on subsequences
with the same attribute values~\cite{Bojanczyk11FO2}. % data words automata.
%
%for the and verification of infinite state systems
%(model checkers: alphabet = long bit-vectors)
%...For the representation of  in model checking, verification and
%
The automata at the core of model checkers
compute on input symbols represented by large bitvectors~\cite{Vardi07ciaa} %\cite{BaierKatoen08MC}
(sets of assignments of Boolean variables) %propositional variables)
and, in practice,  %implementation
each transition accepts a set of such symbols (instead of an individual symbol),
represented by Boolean formula or Binary Decision Diagrams.
%
Following a similar idea, % and generalizing,
in symbolic automata (\SA)~\cite{dAntoniVeanes17CAV,dAntoni21CACM},
transitions are guarded by predicates over infinite  domains.
With appropriate closure conditions on the sets of such predicates, % (alphabet theories),
all the good properties enjoyed over finite alphabets are preserved.

Other extensions of language models  %(automata and grammars)
help in dealing with non-determinism by computing weight values.
With an ambiguous grammar, there may exist several derivations
(\emph{abstract syntax trees} -- AST) % \ie the result of an analyze.
yielding one input word. % structuring a word,
The association of one weight value %in semiring domain
to each AST permits to select a best one (or $n$ bests). % a fixed number of bests
% = ...a ranking of ASTs
%This is roughly the principle of \emph{weighted parsing} approaches
\florent{WARNING: \cite{MorbitzVogler19weighted-parsing} much is more general that weighted parsing}
In \emph{weighted language models}~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing},
like \eg probabilistic context-free grammars % (CFG),
and weighted automata (\WA) \cite{Droste09handbook},
a weight is associated to each transition rule, % production rule
and the rule's weights can be combined with an
associative product operator~$\otimes$ to yield the weight of an AST.
A second operator~$\oplus$
is moreover used to resolve the ambiguity raised by the existence
of several (in general exponentially many) AST
associated to a given input word.
Typically, $\oplus$  selects the best of two weight values.
%Intuitively,~$\oplus$ selects, or ranking, the syntax trees.
The weight domain, equipped with these two operators is, at minima,
a \emph{semiring} %$\Semiring$
where $\oplus$ can be extended to infinite sums,
%\cite{Eilenberg74automata}
such as the Viterbi semiring and the tropical min-plus algebra% % see Figure~\ref{fig:semirings}.
%of domain $\mathbb{R}^+ \cup \{ +\infty\}$,
%where $\oplus$ is min and $\otimes$ is plus .
%by ranking
%making the weight domain a semiring.
%Some efficient specialized parsing algorithms~\cite{Huang05kbest} have been proposed in this context
%% models represented as hypergraphs \cite{Huang05kbest}
%in order to compute the $n$ best syntax trees of a given input word without having to enumerate them all.
%Generally based on dynamic programming, these algorithms rely on
%additional algebraic properties of~$\Semiring$.
%-- see \eg~\cite{Huang05kbest} for some NLP applications.
%The extraction of $n$ best list is useful
%the problem: quantitative parsing or symbolic parsing
%of parsing of words over infinite input alphabet.

In this paper, we present a uniform framework for weighted parsing over infinite input alphabets.
It is based on \emph{symbolic weighted} finite states language models (\swM),
generalizing the Boolean guards of~\SA %in transitions
to functions into an arbitrary semiring,
and generalizing also~\WA, by handling infinite alphabets (see Figure~\ref{fig:hierarchy}).

@@

\philippe{Je ne comprends pas très ce que cette première phrase fait là. Je commencerais
directement le paragraphe avec "This framework relies..."}

@@

A \SWT defines a distance $T(s, t)$ between finite words $s$ and $t$

@@

A  \SWVPA operates sequentially on \emph{nested words}~\cite{AlurMadhusudan09nested},
structured with markup symbols (parentheses), and apt at describing linearizations of trees.
In a parsing context, a \SWVPA $A$ associates a weight value $A(t)$ % \in \Semiring$
to a given nested word $t$, which is itsef the linearization of a weighted AST. %representing a parse tree.

@@

Then, given an input word $s$, the \emph{SW-parsing} problem aims at

@@

$T(s, t) \otimes A(t)$, called the distance between $s$ and $A$ in~\cite{Mohri03EDWA},
\wrt the ranking defined by $\oplus$,

@@

our approach proceeds in two steps.

@@

\begin{example}[Running example]\label{ex:running}
We
illustrate our framework with music transcription: given
a \emph{timeline} of musical events with timestamps from an infinite
domain as input, parse it
into a structured music score. Input events
are pairs $\< \mu, \tau>$, with $\mu \in \Sigma$, where $\Sigma$ stands for the
set of MIDI message symbols~\cite{?}
and  $\tau \in \mathbb{Q}$ is a timestamp. Such inputs typically correspond
to the recording of a live performance. The output of parsing
is a sequence of events (or \emph{notes}) $\< \nu, \tau'>$   in
Common Western Music Notation (CWMN)~\cite{Gould11Notation}
where event symbols $\nu$ belong to the domain $\Delta$
of \emph{pitches} (e.g., A4, G5, etc.), and timestamp $\tau'$
belongs to a rhythmic ``grid'' obtained from recursive divisions:
whole notes ($\musWhole$) splitted in halves ($\musHalf$), halves
in quarters ($\musQuarter$), eights ($\musEighth$), etc.
The following concrete sequences
will be used:
\begin{enumerate}
  \item A performance  $I = [ \<\mu_1, 0.07>, \<\mu_2, 0.72>, \<\mu_3, 0.91>, \<\mu_4, 1.05>, \<\mu_5, 1.36>, \<\mu_6, 1.71>]$,
     over interval $[0,2[$,
  \item A score $O =$ \includegraphics[scale=0.20]{pictures/score5.png}, corresponding to
      a hierarchical structures (measures, tuplets) that can be linearized as a sequence of events
      $[\<\nu_1,0>, \<\nu_2,\frac{3}{4}>, \<\nu_3,\frac{7}{8}>, \<\nu_4,1>, \<\nu_5,\frac{4}{3}>, \<\nu_6\frac{5}{3}>]$
\end{enumerate}
\smallskip

We will show that $O$ is a soution for the
parsing of $I$. Note that several other parsings (e.g.,
\includegraphics[scale=0.20]{pictures/score4.png}) are possible.
SW-parsing associates a weight value
to each solution, and our framework
aims at selecting the best one with respect to this weight.
\endex
\end{example}
